\vspace{-3em} \textbf{ Occultation mapping of Io's surface in the near-infrared \\ I: Inferring static maps}

### 2021-03-02 paper/paper.tex
Robert's comments, adding Robert to the paper

% Title
\title{
    \vspace{-3em}
\textbf{    Occultation mapping of Io's surface in the near-infrared \\ I: Inferring static maps} 
}
% Author list
\author{Fran Bartoli\'c}
\email{fb90@st-andrews.ac.uk}
\affil{Centre for Exoplanet Science, SUPA, School of Physics and Astronomy, University of St. Andrews, St. Andrews, UK}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
\author{Rodrigo Luger}
\author{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
%

@@

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun2002},\citet{rathbun2006} and \citet{rathbun2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots, finding evidence of periodicity in Loki's eruptions and establishing the transient nature of the volcanoes (only Loki is persistently active).
By studying a (spatially resolved) occultation of Io by Europa, \cite{dekleer2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in the near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot1974,saquet2018,morgado2016a} for decades with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.
Understanding the detailed albedo distribution is crucial to constraining the ephemeris to a very high precision.

None of the studies of occultations in the infrared attempted to infer a two-dimensional map of Io's surface.
The light curves are usually fit independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface, with
strong priors on the locations of individual volcanoes based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infrared.
We assume that the map corresponding to a given set of light curves is static (up to an overall amplitude) so we are restricted to modeling the surface during a timespan when the surface emission hasn't changed drastically.
In a subsequent paper we build on the model we have developed here and generalize to a case where the surface is time-variable.

@@

The dataset we use to infer maps of Io's surface consists of 112 sets of observations of Io in the near infrared taken from 1996 until 2018 using different instruments at NASA's IRTF observatory, the NSFCam  1-5 $\mu m$ camera \citep{shure1994} (TODO: WHICH FILTERS), the SpeX 0.8-5.5 $\mu m$ spectrographer/imager \citep{rayner2003} (TODO: DETAILS) and the iShell 1.1-5.3 $\mu m$  spectrographer/imager \citep{johnrayner2016} (TODO: fix citation).
Although the observations span decades, the cadence is non-uniform, there are only a few observations taken between 2008 and 2016.
All of the observations were taken while Io was \emph{in eclipse}, meaning that it was in Jupiter's shadow and all of the observed emission from the surface is due to thermal radiation.
Observations of Io \emph{in sunlight} on the other hand probe both the thermal (volcanic) emission and the surface albedo variations in the near infrared.
Figure~\ref{fig:irtf_max_flux} shows the maximum flux for each occultation light curve in units of $\mathrm{GW}/\mu \mathrm{m}/\mathrm{sr}$ as a function of time.
This value is approximately equal to the total flux emitted from the Jupiter-facing side of Io.
The baseline brightness is varying stochastically by a large amount and there are several notable events of intense volcanic activity.

@@

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected relative position of Jupiter with respect to Io and the orientation of Io at any given time.
To compute the geometry, we use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} and the latest planetary ephemeris DE430 \citep{folkner2014} which is accurate to a few kilometers on Io's surface.
For both Jupiter and Io we need the right ascension and declination and the angular (equatorial) diameter (\textsf{Ang-diam}).
In addition, we need to fix the orientation of Io we need the longitude and the latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}) and the counterclockwise angle between the celestial north pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive  in the direction of west.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 second so we interpolate all values so that we can evaluate them for arbitrary times.

@@

Importantly, conditioned on the parameters specifying the geometry of the occultations being fixed, the \textsf{starry} model is \emph{linear} for both emitted and reflected light maps.
\cite{luger2019} achieved this by representing all rotations, changes of basis transformations and integrals with complicated boundaries which are needed to compute the flux as linear transformations.
Since a sequence of linear mappings is also linear, the predicted flux can be written as
\begin{equation}
    \mathbf{f}=\mathbf{A}\,\mathbf{y}
    \quad,
    \label{eq:linear_model}
\end{equation}
where the column vector $\mathbf{f}$ of shape $(T, 1)$ is the predicted flux for different values of the occultor position and optionally the direction of the illumination source and $\mathbf{A}$ is the design matrix \citep[see appendix B.1. in ][]{luger2021a} of shape $(T, N)$ (with $N=(l+1)^2$) which encodes the operations needed to compute the integrated flux.
If the geometry isn't known precisely the matrix $\mathbf{A}$ is not fixed and the model is in principle no longer linear although one can still use the fact that it is linear when conditioned on a particular values of the nonlinear parameters to speed up inference.

@@

\subsection{The likelihood}
\label{ssec:likelihood}
Assuming we have a single light curve with $T$ data points and a map set by the pixels $\mathbf{p}$, the (Gaussian) log likelihood is given by
\begin{equation}
    \ln\mathcal{L}=-\frac{1}{2}\left[\mathbf{f}_\mathrm{obs}-\mathbf{f} \right]^{\top}
    \boldsymbol{\Sigma}_l^{-1}\left[\mathbf{f}_\mathrm{obs}-\mathbf{f} \right]
    \quad,
    \label{eq:likelihood}
\end{equation}
where $\mathbf{f}_\mathrm{obs}$ is the observed light curve, $\boldsymbol{\Sigma}$ is the data covariance matrix and $\mathbf{f}$ is predicted flux given by
\begin{equation}
    \mathbf{f}=\mathbf{A}'\,\mathbf{p} +b\,\mathbf{1}_T
    \quad,
    \label{eq:flux_model}
\end{equation}
where $\mathbf{A}'\equiv\mathbf{A}\mathbf{P}^\dagger$ and $b$ is a constant flux offset parameter which we have added to account for stray flux which cannot be attributed to Io (it is mostly due to emission from Jupiter).

@@

In order to compute the posterior shrinkage we first compute the design matrices $\mathbf{A}$ using \textsf{starry} for different kinds of events a over the course of a single year: occultations by Jupiter, occultations by Europa, Ganymede and Callisto and phase curves.
We take the ephemeris data from \textsf{JPL HORIZONS} and assume that it is known exactly and we also assume all observations are of the emitted light component independent of whether Io is in sunlight or in eclipse because we are interested in constraining the volcanic emission rather than the albedo.

@@

Fig.~\ref{fig:information_content} shows the posterior shrinkage as a function of $l$ (averaged over all $m$ modes) for phase curve observations (blue lines), occultations by Jupiter (orange), the former two combined (green) and mutual occultations by other Galilean moons (red).
As expected, the mutual occultations of Io by other Galilean moons are by far the most informative with posterior shrinkage of unity at all angular scales considered. 
Occultations by Jupiter are far less informative because we only see one side of Io during an occultation.
Shrinkage for phase curve at odd degrees above $l=2$ is exactly zero because these coefficents are in the nullspace for objects rotating about an axis perpendicular to the line of sight these coefficients and therefore they cannot be constrained using only phase curves \citep{luger2021}. 
Observations of mutual occultations most easily break the degeneracies but they only happen every 6 years and they almost never happen while Io is in eclipse, which means that only the brightest volcanoes are visible above the reflected sunlight.
This is not captured in Figure~\ref{fig:information_content}.

@@

\subsection{Pixels vs. spherical harmonics}
\label{ssec:pixels_vs_harmonics}
We use \textsf{starry} to generate a single simulated light curve of an occultation of Io by Jupiter from an $l=30$ map with known coefficients.
The simulated map consists of a spherical harmonic expansion of a bright spot with a Gaussian profile which we add to a uniform brightness map using the built in \textsf{add\_spot} function in \textsf{starry}.
The expansion is in the quantity $\cos(\Delta\theta)$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere. 
We place the spot at $13^\circ$ N latitude and $51^\circ$ E longitude, we set the diameter of the spot ($2\Delta\theta$) to $5^\circ$ and the amplitude of the spot such that the total \emph{luminosity} of the map increases by 50\% with the addition of the spot.
We generate two light curves with 150 data points, one for the duration of the ingress of the full occultation and the other for the egress.
Because the limb of the occultor sweeps over the disc of Io at different angles during ingress and ingress, this makes it possible break most degeneracies in the map and recover the location of the true spot.
We set the phase of the simulated map to $10^\circ$ W at the beginning of ingress and a phase of $10^\circ$ E at the end of egress.
We assume that the geometry of the occultation is known exactly and set the errorbars such that SNR=50 where SNR is the signal is defined as the maximum flux of light curve.
    We use this dataset to test the difference between setting a prior in the spherical harmonic basis $\mathbf{y}$ and in the pixel basis $\mathbf{p}$ by fitting an $l=20$ map to the dataset.

@@

To ensure that the difference in the inferred maps is not in part due to a difference in the scale of the priors, we take 5000 samples from the prior on $y$ and evaluate $\mathbf{P}\mathbf{y}$ for each; we then compute the standard deviation of these pixels use that as the scale parameter in the exponential prior.
We implement the pixel model in the probabilistic programming language \textsf{numpyro} \citep{phan2019a} which is built on top of the \textsf{JAX} \citep{jax2018github} library, and we fit it using Hamiltonian Monte Carlo with the No-U-Turn-Sampler (NUTS) \citep{hoffman2014}.
\textsf{JAX} is a \textsf{numpy} like library which supports automatic differentiation, parallelization and GPUs.
We run the chains until for 1000 tuning steps and 2000 final steps, monitoring divergences \citep{betancourt2013} and the R-hat  diagnostic  \citep{gelman1992a} to check for convergence.
Since all models in this paper have hundreds if not thousands of parameters, it would not be possible to sample the posterior  without the use of automatic differentiation and Hamiltonian Monte Carlo (at least for non-Gaussian priors).

@@

\subsection{Comparison to a parametric model}
It is instructive to compare the model developed in the previous section to a parametric model in which we assume that the map contains some fixed number of spot and we fit for the positions, amplitudes and the sizes of the spots.
A parametric spot model might be useful if we want to fit for a small number of spots and we know what the map should look like.
To test how the parametrized spot model compares to the model defined in Section~\ref{ssec:fitting_sim_ingress_egress} we fit it to the light curves shown in Figure~\ref{fig:ingress_egress_sim_snr_50}. 
The model consists of 4 parameters: the latitude, longitude, amplitude and size of the spot.
We place uniform priors on the angles and positive Gaussian priors on other parameters. 
We find that if the number of modeled spots matches the number of true spots the model easily converges to the true solution. 
If the number of modeled spots is larger or smaller than the true number the model does not converge due to pathologies in the posterior distribution.

>>>>

% Title
\title{
    \vspace{-3em}
\textbf{    Occultation mapping of Io's surface in the near-infrared \\ I: Inferring static maps} 
}
% Author list
\author{Fran Bartoli\'c}
\email{fb90@st-andrews.ac.uk}
\affil{Centre~for~Exoplanet~Science, SUPA, School~of~Physics~and~Astronomy, University~of~St.~Andrews, St.~Andrews, UK}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
\author{Rodrigo Luger}
\author{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
\author{Robert R. Howell}
\affil{Department~of~Geology~\&~Geophysics, University~of~Wyoming, Laramie, WY}
%

@@

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun2002},\citet{rathbun2006} and \citet{rathbun2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots, finding evidence of periodicity in Loki's eruptions and establishing the transient nature of observed emission from most volcanoes. 
By studying a (spatially resolved) occultation of Io by Europa, \cite{dekleer2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in the near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot1974,saquet2018,morgado2016a} for decades with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.
Understanding the detailed albedo distribution is crucial to constraining the ephemeris to a very high precision.

Most studies of occultations of Io fit multiple light curves independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface (notable exceptions are \cite{spencer1994} and \cite{dekleer2017}), assuming strong priors on the locations of individual volcanoes based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infrared.
We assume that the map corresponding to a given set of light curves is static (up to an overall amplitude) so we are restricted to modeling the surface during a timespan when the surface emission hasn't changed drastically.
In a subsequent paper we build on the model we have developed here and generalize to a case where the surface is time-variable.

@@

The dataset we use to infer maps of Io's surface consists of 112 sets of observations of Io in the near infrared taken from 1996 until 2018 using different instruments at NASA's IRTF observatory, the NSFCam  1-5 $\mu m$ camera \citep{shure1994} (TODO: WHICH FILTERS), the SpeX 0.8-5.5 $\mu m$ spectrographer/imager \citep{rayner2003} (TODO: DETAILS) and the iShell 1.1-5.3 $\mu m$  spectrographer/imager \citep{johnrayner2016} (TODO: fix citation).
Although the observations span decades, the cadence is non-uniform, there are only a few observations taken between 2008 and 2016.
All of the observations were taken while Io was \emph{in eclipse}, meaning that it was in Jupiter's shadow and all of the observed emission from the surface is due to thermal radiation.
Observations of Io \emph{in sunlight} on the other hand probe both the thermal (volcanic) emission and the surface albedo variations in the near infrared.
Figure~\ref{fig:irtf_max_flux} shows the maximum flux for each occultation light curve in units of $\mathrm{GW}/\mu \mathrm{m}/\mathrm{sr}$ as a function of time.
This value is approximately equal to the total flux emitted from the Jupiter-facing side of Io.
The major variation in the baseline brightness is driven by Loki.
Most hotspots can be divided into those who are persistently active for a year or longer at moderate intensity, and those which have bursts of activity and reach very high intensities for short periods of time \cite{dekleer2016}.

@@

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected relative position of Jupiter with respect to Io and the orientation of Io at any given time.
To compute the geometry, we use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} which uses the latest JUPxxx satellite ephemeris \citep{jacobson2015}.
The JUPxxx ephemeris is usually accurate to a few kilometers on Io's surface.
For both Jupiter and Io we need the right ascension and declination and the angular (equatorial) diameter (\textsf{Ang-diam}).
In addition, we need to fix the orientation of Io we need the longitude and the latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}) and the counterclockwise angle between the celestial north pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive  in the direction of west.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 minute, we interpolate these values so that we can evaluate them for arbitrary times.

@@

Importantly, conditioned on the parameters specifying the geometry of the occultations being fixed, the \textsf{starry} model is \emph{linear} for both emitted and reflected light maps.
\cite{luger2019} achieved this by representing all rotations, changes of basis transformations and integrals with complicated boundaries which are needed to compute the flux as linear transformations.
Since a sequence of linear mappings is also linear, the predicted flux can be written as
\begin{equation}
    \mathbf{f}=\mathbf{A}\,\mathbf{y}
    \quad,
    \label{eq:linear_model}
\end{equation}
where the column vector $\mathbf{f}$ of shape $(T, 1)$ is the predicted flux for different values of the occultor position and $\mathbf{A}$ is the design matrix \citep[see appendix B.1. in ][]{luger2021a} of shape $(T, N)$ (with $N=(l+1)^2$) which encodes the operations needed to compute the integrated flux at different viewing angles of the occulted sphere.
If the geometry isn't known precisely the matrix $\mathbf{A}$ is not fixed and the model is in principle no longer linear although one can still use the fact that it is linear when conditioned on a particular values of the nonlinear parameters to speed up inference.

@@

\subsection{The likelihood}
\label{ssec:likelihood}
Assuming we have a single light curve with $T$ data points and a map set by the pixels $\mathbf{p}$, the (Gaussian) log likelihood is given by
\begin{equation}
    \ln\mathcal{L}=-\frac{1}{2}\left[\mathbf{f}_\mathrm{obs}-\mathbf{f} \right]^{\top}
    \boldsymbol{\Sigma}_l^{-1}\left[\mathbf{f}_\mathrm{obs}-\mathbf{f} \right]
    \quad,
    \label{eq:likelihood}
\end{equation}
where $\mathbf{f}_\mathrm{obs}$ is the observed light curve, $\boldsymbol{\Sigma}$ is the data covariance matrix and $\mathbf{f}$ is predicted flux given by
\begin{equation}
    \mathbf{f}=\mathbf{A}'\,\mathbf{p} +b\,\mathbf{1}_T
    \quad,
    \label{eq:flux_model}
\end{equation}
where $\mathbf{A}'\equiv\mathbf{A}\mathbf{P}^\dagger$ and $b$ is a constant flux offset parameter which we have added to account for stray flux which cannot be attributed to Io and is likely due to residual light from Jupiter.

@@

In order to compute the posterior shrinkage we first compute the design matrices $\mathbf{A}$ using \textsf{starry} for different kinds of observations of Io in the period starting on the 1st of January 2009 and ending on the 1st of May 2010.
We chose this period because it covers the full season of mutual occultations which occur every 6 years.
We compute $\mathbf{A}$ for all observable Io by Galilean moons during that period, for occultations of Io by Jupiter and phase curves.
We make sure that the number of observations and the cadence for each observation covers all possible geometries because the purpose of this exercise is do determine the upper bound on what we can learn about the surface. 
We take the ephemeris from \textsf{JPL HORIZONS} and assume that it is known exactly; we also assume all observations are of the emitted light component independent of whether Io is in sunlight or in eclipse because we are interested in constraining the volcanic emission rather than the albedo.

@@

Fig.~\ref{fig:information_content} shows the posterior shrinkage as a function of $l$ (averaged over all $m$ modes) for phase curve observations (blue lines), occultations by Jupiter (orange), the former two combined (green) and mutual occultations by other Galilean moons (red).
As expected, the mutual occultations of Io by other Galilean moons are by far the most informative with posterior shrinkage of unity at all angular scales considered. 
Occultations by Jupiter are far less informative because we only see one side of Io during an occultation.
Shrinkage for phase curve at odd degrees above $l=2$ is exactly zero because these coefficents are in the nullspace for objects rotating about an axis perpendicular to the line of sight these coefficients and therefore they cannot be constrained using only phase curves \citep{luger2021}. 
Observations of mutual occultations most easily break the degeneracies but the drawback of these types of observations is that they only happen every 6 years and they almost never happen while Io is in eclipse, meaning that only the brightest volcanoes are visible above the reflected sunlight.
This fact is not captured in Figure~\ref{fig:information_content}.

@@

\subsection{Pixels vs. spherical harmonics}
\label{ssec:pixels_vs_harmonics}
We use \textsf{starry} to generate a single simulated light curve of an occultation of Io by Jupiter from an $l=30$ map with known coefficients.
The simulated map consists of a spherical harmonic expansion of a bright spot with a Gaussian profile which we add to a uniform brightness map using the built in \textsf{add\_spot} function in \textsf{starry}.
The expansion is in the quantity $\cos(\Delta\theta)$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere. 
We place the spot at $13^\circ$ N latitude and $51^\circ$ E longitude, we set the diameter of the spot ($2\Delta\theta$) to $5^\circ$ and the amplitude of the spot such that the total \emph{luminosity} of the map increases by 50\% with the addition of the spot.
We generate two light curves with 150 data points, one for the duration of the ingress of the full occultation and the other for the egress.
Because the limb of the occultor sweeps over the disc of Io at different angles during ingress and ingress, this makes it possible break most degeneracies in the map and recover the location of the true spot.
We set the phase of the simulated map to be $10^\circ$ E at the beginning of ingress and $10^\circ$ W at the end of egress.
We assume that the geometry of the occultation is known exactly and set the errorbars such that SNR=50 where SNR is the signal is defined as the maximum flux of light curve.
    We use this dataset to test the difference between setting a prior in the spherical harmonic basis $\mathbf{y}$ and in the pixel basis $\mathbf{p}$ by fitting an $l=20$ map to the dataset.

@@

To ensure that the difference in the inferred maps is not in part due to a difference in the scale of the priors, we take 5000 samples from the prior on $y$ and evaluate $\mathbf{P}\mathbf{y}$ for each; we then compute the standard deviation of these pixels use that as the scale parameter in the exponential prior.
We implement the pixel model in the probabilistic programming language \textsf{numpyro} \citep{phan2019a} which is built on top of the \textsf{JAX} \citep{jax2018github} library, and we fit it using Hamiltonian Monte Carlo with the No-U-Turn-Sampler (NUTS) \citep{hoffman2014}.
\textsf{JAX} is a \textsf{numpy} like library which supports automatic differentiation, parallelization and GPUs.
We run the chains for 1000 tuning steps and 2000 final steps, monitoring divergences \citep{betancourt2013} and the R-hat  diagnostic  \citep{gelman1992a} to check for convergence.
Since all models in this paper have hundreds if not thousands of parameters, it would not be possible to sample the posterior  without the use of automatic differentiation and Hamiltonian Monte Carlo (at least for non-Gaussian priors).

@@

\subsection{Comparison to a parametric model}
It is instructive to compare the model developed in the previous section to a parametric model in which we assume that the map contains some fixed number of spots and we fit for the positions, amplitudes and the sizes of the spots.
A parametric spot model might be useful if we want to fit for a small number of spots and we know what the map should look like.
To test how the parametrized spot model compares to the model defined in Section~\ref{ssec:fitting_sim_ingress_egress} we fit it to the light curves shown in Figure~\ref{fig:ingress_egress_sim_snr_50}. 
The model consists of 4 parameters: the latitude, longitude, amplitude and size of the spot.
We place uniform priors on the angles and positive Gaussian priors on other parameters. 
We find that if the number of modeled spots matches the number of true spots the model easily converges to the true solution. 
If the number of modeled spots is larger or smaller than the true number the model does not converge due to pathologies in the posterior distribution.
